---
title: "Project - COVID19"
format:   
  html: 
    toc: true
    toc-depth: 3
    standalone: true
    embed-resources: true
    code-fold: true
    number-sections: true
editor_options: 
 
---

```{r}
#| label: packages-data-functions
#| message: false
library(tidyverse)
library(patchwork)            
library(tidymodels)
library(readxl)
library(glue)
library(corrplot)
library(dplyr)
library(knitr)
library(tidyr)
WDI <- read_csv("~/Documents/Constructor studies/DS concepts and DT tools/github_repositories/R/HW#3 COVID/Project_COVID19_aaldaniyaz/data/WDI.csv")
WHO <- read_csv("~/Documents/Constructor studies/DS concepts and DT tools/github_repositories/R/HW#3 COVID/Project_COVID19_aaldaniyaz/data/WHO-COVID-19.csv")
```


# COVID-19 time evolution {#sec-1}


## Covid Cases and Deaths in Germany's 1st wave {#sec-1.1}

```{r}
WHO |> filter(Country=="Germany", Date_reported<="2020-08-31")|>
    pivot_longer(cols = c(New_cases, New_deaths), names_to = "Case_Death", values_to = "Count")|>
    ggplot(aes(x = Date_reported, y = Count, color = Case_Death)) +
  geom_line(aes(group=Case_Death))+
      labs(x = "",
       y = "Number of Cases",
       title = "The number of new and death cases through time",
       color="")+
      scale_color_manual(values = c("New_cases" = "blue", "New_deaths" = "green"))+
      theme_minimal()+
  scale_y_continuous(breaks = seq(0, max(WHO$New_cases), by = 500))+
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month")
```


## Cumulative cases in Germany, Italy, France, and UK {#sec-1.2}
```{r}
WHO|>
  filter(Country %in% c("Germany", "Italy", "France", "The United Kingdom"), Date_reported<="2020-08-31")|>
  group_by(Country)|>
  arrange(Country, Date_reported)|>
  mutate(C_cases=cumsum(New_cases), C_deaths=cumsum(New_deaths))|>
  select(Country, Date_reported, C_cases, C_deaths)|>
  pivot_longer(cols = c("C_cases", "C_deaths"), names_to = "Cases_deaths", values_to = "Count")|>
  ggplot(aes(x=Date_reported, y=Count, color=Country))+
  geom_line(aes(group=Country))+
  labs(x = "",
       y = "Number of cases",
       title = "The number of cumulative cases and deaths through time")+
  facet_wrap(~ Cases_deaths,scales = "free_y",labeller = labeller(Cases_deaths = c("C_cases" = "Cumulative Cases", "C_deaths" = "Cumulative Deaths")))+
  scale_x_date(date_labels = "%b", date_breaks = "2 month")+
  theme_minimal()
```


## Smoothing daily data {#sec-1.3}
```{r}
smooth3 <- function(x) {
  (x + lag(x, n = 1, default = 0) + lag(x, n = 2, default = 0)) / 3
}

smooth7 <- function(x) {
  (x + lag(x, n = 1, default = 0) + lag(x, n = 2, default = 0) + lag(x, n = 3, default = 0) +
     lag(x, n = 4, default = 0) + lag(x, n = 5, default = 0) + lag(x, n = 6, default = 0)) / 7
}

smooth10 <- function(x) {
  (x + lag(x, n = 1, default = 0) + lag(x, n = 2, default = 0) + lag(x, n = 3, default = 0) +
     lag(x, n = 4, default = 0) + lag(x, n = 5, default = 0) + lag(x, n = 6, default = 0) +
     lag(x, n = 7, default = 0) + lag(x, n = 8, default = 0) + lag(x, n = 9, default = 0)) / 10
}

who_smoothed <- WHO|>
  group_by(Country)|>
  mutate(
    New_cases_smooth3 = smooth3(New_cases),
    New_cases_smooth7 = smooth7(New_cases),
    New_cases_smooth10 = smooth10(New_cases),
    New_deaths_smooth3 = smooth3(New_deaths),
    New_deaths_smooth7 = smooth7(New_deaths),
    New_deaths_smooth10 = smooth10(New_deaths)
  )

who_smoothed|>
  filter(Country == "Germany", Date_reported <= "2020-08-31") |>
  pivot_longer(
    cols = c("New_cases", "New_cases_smooth3", "New_cases_smooth7", "New_cases_smooth10"),
    names_to = "Smooth_Variable",
    values_to = "Count"
  )|>
  ggplot(aes(x = Date_reported, y = Count, color = Smooth_Variable)) +
  geom_line(aes(group = Smooth_Variable)) +
  labs(
    x = "",
    y = "Number of Cases",
    title = "Comparison of Smoothed Cases and Daily New Cases in Germany until 2020-08-31",
    color = ""
  ) +
  theme_minimal() +
  facet_wrap(~ Smooth_Variable, scales = "free_y", labeller = labeller(Smooth_Variable = c("New_cases" = "New cases through time","New_cases_smooth3" = "Smoothed Cases for the last 3 days", "New_cases_smooth7" = "Smoothed Cases for the last 7 days", "New_cases_smooth10" = "Smoothed Cases for the last 10 days"))) +
  theme(legend.position = "none")
```


## How do deaths follow cases? {#sec-1.4}
```{r}
shiftscale <- function(x, shift, scale) {
  lagged_x <- lag(x, shift, default = 0)
  scaled_x <- lagged_x * scale
  return(scaled_x)}

who_smoothed|>
  mutate(shiftscale_cases = shiftscale(New_cases_smooth7, 5, 0.062))|>
  filter(Country == "Germany", Date_reported <= "2020-08-31")|>
  pivot_longer(cols = c("shiftscale_cases", "New_deaths_smooth7"),
               names_to = "Variable",
               values_to = "Count")|>
  ggplot(aes(x = Date_reported, y = Count, color = Variable)) +
  geom_line(aes(group = Variable)) +
  labs(x = "",
       y = "Number of Cases/Deaths",
       title = "Comparison of Shiftscaled Cases and Deaths for Germany till 2020-08-31",
       subtitle = "(the time series are smoothed over the last 7 days)",
       color = "")+
  scale_color_manual(values = c("New_deaths_smooth7" = "red","shiftscale_cases" = "blue"),
                     labels = c("Deaths (Smoothed)", "Scaled and Shifted New Cases (Smoothed)"))+
  scale_x_date(date_labels = "%b %Y", date_breaks = "2 month")+
  theme_minimal()
```


## When does the wave break? {#sec-1.5}
```{r}
who_smoothed|>
  mutate(shiftscale_cases = shiftscale(New_cases_smooth7, 6, 0.035))|>
  filter(Country == "Germany", Date_reported >= "2020-07-01", Date_reported <= "2020-12-31")|>
  pivot_longer(cols = c("shiftscale_cases", "New_deaths_smooth7"),
               names_to = "Variable",
               values_to = "Count")|>
  ggplot(aes(x = Date_reported, y = Count, color = Variable)) +
  geom_line(aes(group = Variable)) +
  labs(x = "",
       y = "Number of Cases/Deaths",
       title = "Comparison of Shiftscaled Cases and Deaths for Germany",
       subtitle = "from 2020-07-01 till 2020-12-31(the time series are smoothed over the last 7 days)",
       color = "")+
  scale_color_manual(values = c("New_deaths_smooth7" = "red","shiftscale_cases" = "blue"),
                     labels = c("Deaths (Smoothed)", "Scaled and Shifted New Cases (Smoothed)"))+
  scale_x_date(date_labels = "%b %Y", date_breaks = "2 month")+
  theme_minimal()
```



# World Development Indicators {#sec-2}
```{r}
who_smoothed |>
  mutate(Diff_cases_smooth7 = New_cases_smooth7 - lag(New_cases_smooth7, default = 0)) |>
  filter(Country == "Germany", Date_reported >= "2020-03-01", Date_reported <= "2020-04-15") |>
  pivot_longer(cols = c("New_cases_smooth7", "Diff_cases_smooth7"),
               names_to = "Variable",
               values_to = "Count") |>
  ggplot(aes(x = Date_reported, y = Count, color = Variable)) +
  geom_line(aes(group = Variable)) +
  labs(x = "",
       y = "Number of Cases",
       title = "Smoothed New Cases and Derivative (Diff) in Germany",
       subtitle = "from 1st of March to 15th of April (the time series are smoothed over the last 7 days)",
       color = "Variable") +
  theme_minimal() +
  facet_grid(Variable ~ ., scales = "free_y", labeller = labeller(Variable = c("Diff_cases_smooth7" = "Derivative of new cases", "New_cases_smooth7" = "New cases (smoothed)"))) +
  scale_x_date(date_breaks = "2 days", date_labels = "%d") +
  theme(legend.position = "none")
```


```{r}
## Correlation {#sec-2.1}
wdi <- WDI |>
  rename(
    urban_pop = `Urban population (% of total population)`,
    rural_pop = `Rural population (% of total population)`,
    pop_lower_half_median = `Proportion of people living below 50 percent of median income (%)`,
    pop = `Population, total`,
    pop_older65 = `Population ages 65 and above (% of total population)`,
    pop_density = `Population density (people per sq. km of land area)`,
    physicians_per_1000 = `Physicians (per 1,000 people)`,
    life_expectancy = `Life expectancy at birth, total (years)`,
    gdp_per_capita = `GDP per capita, PPP (current international $)`
  )
```


```{r}
cor_matrix <- cor(wdi[, sapply(wdi, is.numeric)], use = "complete.obs")

corrplot(cor_matrix, tl.cex = 0.7, tl.col = "black", tl.srt = 45)
```

# PCA {#sec-2.2}
```{r}
# Create wdi_noNA_small by removing all rows with any NA's
wdi_noNA_small <- na.omit(wdi)

# Find the two variables with the most NA's
na_count <- colSums(is.na(wdi))
top_na_vars <- names(sort(na_count, decreasing = TRUE)[1:2])

# Remove the two variables with the most NA's
wdi_noNA_large <- wdi %>%
  select(-all_of(top_na_vars)) %>%
  na.omit()

```


```{r}
missing_countries_small <- anti_join(wdi, wdi_noNA_small, by = "iso3c") %>%
  filter(pop > 1e6)

kable(missing_countries_small[, c("iso3c", "country", "pop")], 
      caption = "Countries Missing in wdi_noNA_small with Population > 1 Million")

```


```{r}
# Create wdi_noNA_small by removing all rows with any NA's
wdi_noNA_small <- na.omit(wdi)

# Find the two variables with the most NA's
na_count <- colSums(is.na(wdi))
top_na_vars <- names(sort(na_count, decreasing = TRUE)[1:2])

# Remove the two variables with the most NA's
wdi_noNA_large <- wdi %>%
  select(-all_of(top_na_vars)) %>%
  na.omit()
```


```{r}
missing_countries_small <- anti_join(wdi, wdi_noNA_small, by = "iso3c") %>%
  filter(pop > 1e6)

kable(missing_countries_small[, c("iso3c", "country", "pop")], 
      caption = "Countries Missing in wdi_noNA_small with Population > 1 Million")

```



PCA SMALL
```{r}
# Scale the numeric variables for PCA
wdi_small_pca <- prcomp(select(wdi_noNA_small, where(is.numeric)), scale. = TRUE)

```


```{r}
explained_variance_small <- summary(wdi_small_pca)$importance[2,]

# Plot explained variance
qplot(x = 1:length(explained_variance_small), 
      y = explained_variance_small, 
      geom = "line") +
  geom_point() +
  labs(title = "Explained Variance (wdi_noNA_small)", 
       x = "Principal Component", 
       y = "Proportion of Variance Explained")

```
The explained variance plot shows:

PC1 explains nearly 50% of the variance.
PC2 explains about 15% of the variance.
The remaining principal components explain progressively smaller amounts of variance.
This suggests that the first two principal components capture a significant portion of the variability in the data.

```{r}
kable(wdi_small_pca$rotation[, 1:4], 
      caption = "First Four Principal Components (wdi_noNA_small)")

```


```{r}
pc_small_df <- data.frame(wdi_noNA_small, wdi_small_pca$x)

ggplot(pc_small_df, aes(x = PC1, y = PC2, color = region, label = iso3c)) +
  geom_text(size = 3) +
  labs(title = "Countries in PC1 and PC2 (wdi_noNA_small)", 
       x = "PC1", y = "PC2")

```

This analysis suggests that PC1 captures economic development, while PC2 might reflect demographic or regional-specific variations.


#PCA Large
```{r}
# Scale the numeric variables for PCA
wdi_large_pca <- prcomp(select(wdi_noNA_large, where(is.numeric)), scale. = TRUE)
```


```{r}
explained_variance_large <- summary(wdi_large_pca)$importance[2,]

# Plot explained variance
qplot(x = 1:length(explained_variance_large), 
      y = explained_variance_large, 
      geom = "line") +
  geom_point() +
  labs(title = "Explained Variance (wdi_noNA_large)", 
       x = "Principal Component", 
       y = "Proportion of Variance Explained")

```
The explained variance plot shows:

PC1 explains approximately 50% of the variance.
PC2 explains around 18% of the variance.
The remaining components explain smaller portions of the variance.

```{r}
kable(wdi_large_pca$rotation[, 1:4], 
      caption = "First Four Principal Components (wdi_noNA_large)")

```


```{r}
pc_large_df <- data.frame(wdi_noNA_large, wdi_large_pca$x)

ggplot(pc_large_df, aes(x = PC1, y = PC2, color = region, label = iso3c)) +
  geom_text(size = 3) +
  labs(title = "Countries in PC1 and PC2 (wdi_noNA_large)", 
       x = "PC1", y = "PC2")

```
PC1 seems to separate countries based on economic or development factors.
PC2 might capture differences related to geographic or demographic factors

```{r}
